Communicative reduction in referring expressions within a multi-player negotiation game

### 2023-02-01 write-ups/cogsci23_paper/manuscript.Rmd
final!

---
title: "Communicative reduction in referring expressions within a multi-player negotation game"
bibliography: [ aaflowers.bib]
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

The formation of ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986; @krauss1964changes]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize within each pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs. This task has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, and when they change [@brennan1996conceptual; @metzing2003conceptual; @leung2020parents; @weber2003; @yoonAudienceDesignMultiparty2019].

The implicit theoretical claim is that reduction to partner-specific conventions occurs *whenever* people interact repeatedly in ways that require reference to some objects without adequate conventional names. But do these phenomena generalize beyond artificial reference game environments?

@@

We build on work by @mankewitz2021 who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datasets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

Our work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge are able to develop conventionalized reference expressions. We explore whether the phenomena found in classic dyadic iterated reference games extend to this more naturalistic domain, and whether the results differ across the individual and shared incentive conditions. The key questions we addressed were:

1. Did the referring expressions reduce over time?

@@

3. How did the different incentive conditions influence the referring expressions?

Historically, studies of convention formation have had to rely on proxies such as reduction, subjective judgments of similarity, and manual measures of lexical overlap. In contrast, we use natural language processing tools for mapping sentences to semantic vector spaces to quantify semantic similarities between sentences, following recent work by @hawkins2020 and @boyce2022.

@@

For pairwise comparisons between referring expressions produced during the game, we smoothed out the random sampling and sparseness of which flowers were presented on the same trial by comparing descriptions with those from the same trial and the two later trials (thus the similarity at trial 10 comes from 10-10, 10-11 and 10-12 comparisons).

@@

We address each of the questions of interest listed at the end of the Introduction in order.

@@

In each trial of the game, there were 6 flowers visible on the screen, and participants wanted to each pick a different one. In general, each trial contained references to 2 or 3 distinct flowers (individual utilities: mean of `r  mean_flowers |> pluck(2,1)`, shared utilities: `r mean_flowers |> pluck(2,2)`). Most players referred to one flower each trial (individual utilities: `r  mean_person |> pluck(2,1)`, shared utilities: `r mean_person |> pluck(2,2)`).

@@

For descriptions of different flowers, cosine similarity decreased over time (trial: `r stats_text(wi_g_bet_f,6)`, see Figure \ref{during} lower left and middle panels). Utterances being produced by the same person did not have an effect (`r stats_text(wi_g_bet_f,8)`) or an interaction with trial number (`r stats_text(wi_g_bet_f,7)`) .

These results replicate and generalize the pattern found in the prior literature [@hawkins2020].

@@

```
### Between games

@@

For descriptions of the same flower across games, we found that cosine similarities decreased over time  (trial: `r stats_text(bet_g_wi_f,4)`, see Figure \ref{during} upper right panel). This decrease is consistent with the predictions.

@@

Referring descriptions later in a game were more similar to the convention than descriptions earlier in the game  (trial: `r stats_text(to_e_wi_f,6)`, see Figure \ref{toend} upper panels). Utterances were more similar to the convention given by the same person (versus the convention given by a groupmate) (`r stats_text(to_e_wi_f,8)`), and this increased over time (`r stats_text(to_e_wi_f,7)`). This is consistent with the expected pattern of convention formation.

@@

```{r withinend, fig.env="figure", fig.pos = "h", fig.align = "center", out.width="90%", fig.width=4, fig.height=2, fig.cap = "Cosine similarities between flower descriptions provided by participants after the game. \\label{withinend}"}
in_end|> 
  filter(player2!=player1) |> 
  mutate(distance=case_when(
    game1==game2 ~ "Same group",
    T ~ "Different group"),
    same_flower=ifelse(ref1==ref2, "Same flower", "Different flower"))|> 
ggplot( aes(x=str_c(same_flower,"\n", distance),y=sim,color=condition))+
  labs(y="Cosine Similarity", x=NULL)+
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width=.3))+theme(legend.position=c(0.2,.75), legend.title=element_blank(), legend.background=element_blank())+
        ggthemes::scale_color_solarized()
```
## Individual and shared utility conditions

@@

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found some modest condition differences within games. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition.

>>>>

---
title: "Communicative reduction in referring expressions within a multi-player negotiation game"
bibliography: [ aaflowers.bib]
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

The formation of ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986; @krauss1964changes]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize within each pair, as pairs tend to stick to the same description, but nicknames are idiosyncratic and differ between pairs. This task has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, and when they change [@brennan1996conceptual; @metzing2003conceptual; @leung2020parents; @weber2003; @yoonAudienceDesignMultiparty2019].

The implicit theoretical claim is that reduction to partner-specific conventions occurs *whenever* people interact repeatedly in ways that require reference to some objects without adequate conventional names. Indeed, reduction does occur regardless of modality, having been found in both oral and written communication [@brennan1996conceptual; @hawkins2020]. But do these phenomena generalize beyond artificial reference game environments?

@@

We build on work by @mankewitz2021, who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datasets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

Our work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge are able to develop conventionalized reference expressions. We explore whether the phenomena found in classic dyadic iterated reference games can also be found in this more naturalistic domain, and whether the results differ across the individual and shared incentive conditions. The key questions we addressed were:

1. Did referring expressions reduce over time?

@@

3. How did different incentive conditions influence referring expression formation?

Historically, studies of convention formation have had to rely on proxies for convention formation such as reduction, subjective judgments of similarity, and manual measures of lexical overlap. In contrast, we use natural language processing tools for mapping sentences to semantic vector spaces to quantify semantic similarities between sentences, following recent work by @hawkins2020 and @boyce2022.

@@

For pairwise comparisons between referring expressions produced during the game, we smoothed out the random sampling and sparseness of which flowers were presented on the same trial by comparing descriptions with those from the same trial and the two later trials (thus the similarity at trial 10 comes from comparisons of trial 10 to trials 10, 11, and 12).

@@

We address each of the questions of interest in order: 1) did referring expressions reduce over time, 2) how referring expressions changed during games, and 3) how the incentive manipulation influenced referring expressions.

@@

In each trial of the game, there were 6 flowers visible on the screen, and participants wanted to each pick a different one. In general, each trial contained references to 2 or 3 distinct flowers: individual utilities: mean of `r  mean_flowers |> pluck(2,1)`, shared utilities: `r mean_flowers |> pluck(2,2)`. Most players referred to one flower each trial: individual utilities: `r  mean_person |> pluck(2,1)`, shared utilities: `r mean_person |> pluck(2,2)`.

@@

@@

For descriptions of different flowers, cosine similarity decreased over time (trial: `r stats_text(wi_g_bet_f,6)`, see Figure \ref{during} lower left and middle panels). Utterances being produced by the same person did not have an effect (`r stats_text(wi_g_bet_f,8)`) or an interaction with trial number (`r stats_text(wi_g_bet_f,7)`).

These results replicate and generalize the pattern found in prior literature: convergence in expressions for the same referent  games and divergence between referents [@hawkins2020].

### Between games

@@

For descriptions of the same flower across games, we found that cosine similarities decreased over time  (trial: `r stats_text(bet_g_wi_f,4)`, see Figure \ref{during} upper right panel). This between-game decrease is also consistent with prior reports in more restricted reference games.

@@

Referring descriptions later in a game were more similar to the convention than descriptions earlier in the game  (trial: `r stats_text(to_e_wi_f,6)`, see Figure \ref{toend} upper panels). Utterances were more similar to the convention given by the same person (versus the convention given by a groupmate) (`r stats_text(to_e_wi_f,8)`), and this increased over time (`r stats_text(to_e_wi_f,7)`). This finding is consistent with the expected pattern of convention formation.

@@

```{r withinend, fig.env="figure", fig.pos = "h", fig.align = "center", out.width="90%", fig.width=4, fig.height=2, fig.cap = "Cosine similarities between flower descriptions provided by participants after the game. Error bars are 95\\% CIs, but some are hidden by the point size. \\label{withinend}"}
in_end|> 
  filter(player2!=player1) |> 
  mutate(distance=case_when(
    game1==game2 ~ "Same group",
    T ~ "Different group"),
    same_flower=ifelse(ref1==ref2, "Same flower", "Different flower"))|> 
ggplot( aes(x=str_c(same_flower,"\n", distance),y=sim,color=condition))+
  labs(y="Cosine Similarity", x=NULL)+
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width=.3))+theme(legend.position=c(0.2,.75), legend.title=element_blank(), legend.background=element_blank())+
        ggthemes::scale_color_solarized()
```
## Individual and shared utility conditions

@@

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found some modest condition differences within games. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition, and the modest magnitude of the effect is consistent with the relatively subtle nature of this manipulation, which did not have a large effect on choice behavior [@mankewitz2021].
